Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-16 ebayes_rnaseq_heterosis.tex
self-contained
through other methods

\graphicspath{{../temp/rna-seq-stan/fig-rm-stan_laplace/}}

@@

As defined in Section \ref{s:heterosis}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
\label{e:hypotheses}
H_0:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_1:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\end{align}

\section{Simulation study}
\label{s:simulation}

\subsection{Coverage for our model}

\input{../temp/analyze-heterosis/FIGURES/coverage}

\subsection{Heterosis detection}

@@

\section{}
\subsection{Stan model for heterosis}

>>>>

\graphicspath{{include/}}

@@

As defined in Section \ref{s:heterosis}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
\label{e:hypotheses}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a sorted list of genes potentially involved in phenotypic heterosis.

\section{Simulation study based on maize experiment}
\label{s:simulation}

% \subsection{Coverage for our model}
% \input{../temp/analyze-heterosis/FIGURES/coverage}

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}. We compared our approach to approaches using {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.

\subsection{Construct simulated data}

Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety. We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one which resulted in 27,888 genes remaining.

To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means. We estimated sample-specific normalization factors to account for sequencing depth. Then, we estimated common a common dispersion for each gene, but shared across the samples. Finally, we obtained independent estimates for the gene-variety specific location parameters. The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study. A gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have low parent heterosis while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have high parent heterosis.

Using these parameters and normalization factors, we simulated data according to the negative-binomial model in equation \eqref{e:data} independently for each gene. For each simulation, low count genes, i.e. those with three or more zeros in any variety or whose mean count across the varieties was less than one, were removed. A random subset of 25,000 genes was retained from those with high enough counts. This simulation process was repeated 10 times for each of 4, 8, and 16 replicates per variety (reusing normalization factors for the higher number of replicates).

\subsection{Alternative methods}

Existing methods are not designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}. and {\tt ShrinkBayes} to provide a statistic representing the plausibility of heterosis.

The approach used by {\tt edgeR} naturally provides testing of hypotheses for differential expression amongst two varieties \citep{robinson2007moderated, robinson2010edgeR}. If the hybrid's mean expression does not differ from a parent, then the hybrid does not exhibit heterosis. We can calculate two pvalues: one for the null hypothesis that the hybrid's mean expression is the same as parent 1 and the second for the null hypothesis that the hybrid's mean expression is the same as parent 2. But, even if both pvalues are small, the hybrid's mean expression could be between the two parents, i.e. mid-parent heterosis, and therefore not be exhibiting extreme heterosis. We construct a statistic that is 0 if the estimate of mean expression 
for the hybrid is between the estimates for the parents. Otherwise, the statistic is one minus the pvalue above corresponding to whichever parent's mean estimate is closer to that of the hybrid.

The {\tt baySeq} approach allows for a wider range of hypotheses including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct \citep{hardcastle2010bayseq, hardcastle2012baySeq}. Hypotheses $H_2$ and $H_5$ are the only two that indicate the possibility of heterosis. Similar to the {\tt edgeR} approach, we construct a statistic that is 0 if the estimated hybrid mean is between the parental means and otherwise is the sum of the posterior probabilities for hypotheses $H_2$ and $H_5$.

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation with INLA is that it only provides approximations to marginal posteriors rather than the full posterior. Thus we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})$. To circumvent this limitation, we reparameterized following \cite{ji2014estimation}. Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. We implemented this model using {\tt ShrinkBayes} and constructed a statistic based on posterior median of these parameters. Specifically, the statistic is 0 if $\left|\hat{\delta}_f  \right| \le \left| \hat{\alpha}_f \right|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.

\subsection{Results}

@@

\section{}

%\input{stan_models}
